---
title: "Australian's age and sex"
subtitle: Data from the 2021 national census
date: 2023-02-03
date-format: long
author: Matthew Andreotta
format:
    html:
        theme:  flatly
echo:   false
include:    false
warning:    false
---

This webpage shares results from the the 2021 Australian census on age and sex.
This data may be useful for building survey samples representative of Australians.
In particular, @tbl-age-decades describes the proportion of Australians across age categories and @tbl-sex describes the proportion of Australians across sexes.

Raw count data is provided in @tbl-counts.

# Introduction

In 2021, the Australian Bureau of Statistics (ABS) conducted a census of Australians.
Methodology can be found on the [ABS' website](https://www.abs.gov.au/census/guide-census-data/census-methodology/2021).
For age, respondents were asked the person's date of birth or age.
For sex, respondents were asked to select whether the person was: "male", "female", or "non-binary sex".[^longnote]
In this document, I have used the same terminology as the census.

[^longnote]: **A word of warning**. As per ABS' [own words](https://www.abs.gov.au/articles/analysis-non-binary-sex-responses#2021-census-results):

    > The non-binary sex option was marked on the Census form for 43,220 respondents or 0.17% of the Australian population. Analysis indicates this number is not indicative of any single characteristic. Responses show the concept of non-binary sex was not consistently understood and was perceived in different ways by different people. Results cannot be used as a measure of gender diversity, non-binary genders or trans populations. Additionally, it cannot be used as a measure for diverse sexualities, nor can it be interpreted as the number of people with variations of sex characteristics.

    For these reasons and more, I have ignored non-binary sex responses when presenting sex statistics.

    I believe the age data used for the statistics in this document do not include respondents who selected the non-binary sex response.

```{r load-packages}
library(tidyverse) # data wrangling
library(gt) # tables
knitr::opts_knit$set(root.dir = '..') # change working directory when render
```


```{r helper-functions}
specify_decimal <- function(number_to_round, number_of_decimals = 2) {
    format(
        round(number_to_round, number_of_decimals),
        nsmall = number_of_decimals
    )
}
```

```{r load-data}
data <- read_csv(
        file = "data/2021_sex-and-age.csv",
        col_names = c("age", "male", "female", "total"),
        skip = 1,
    )
```

# Proportions

## Age

```{r tbl-age-decades}
#| include: true
#| tbl-cap: The total count and proportion of Australian's age, categorised in decade-wide age groups.
#| label: tbl-age-decades
data |>
    mutate(age_group = case_when(
        age < 30                ~ "18-29",
        age >= 30 & age < 40    ~ "30-40",
        age >= 40 & age < 50    ~ "40-50",
        age >= 50 & age < 60    ~ "50-60",
        age >= 60 & age < 70    ~ "60-70",
        age >= 70               ~ "70+",
    )) |>
    group_by(age_group) |>
    summarise(total = sum(total)) |>
    ungroup() |>
    mutate(percentage = specify_decimal(total / sum(total) * 100)) |>
    select(age_group, total, percentage) |>
    rename_with(str_to_sentence) |>
    gt() |>
    tab_style(
        locations = cells_column_labels(columns = everything()),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_style(
        locations = cells_body(columns = "Age_group"),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_options(table.border.top.style = "hidden") |>
    cols_label(
        Age_group = "Age (Years)",
        Percentage = "Proportion (%)"    
    )
```

## Sex

```{r tbl-sex}
#| include: true
#| tbl-cap: The total count and proportion of Australian's sex.
#| label: tbl-sex
data |>
    select(-age, -total) |>
    summarise(across(everything(), sum)) |>
    pivot_longer(
        everything(),
        names_to = "sex",
        values_to = "total"
    ) |>
    mutate(sex = str_to_sentence(sex)) |>
    mutate(percentage = specify_decimal(total / sum(total) * 100)) |>
    select(sex, total, percentage) |>
    arrange(desc(total)) |>
    rename_with(str_to_sentence) |>
    gt() |>
    tab_style(
        locations = cells_column_labels(columns = everything()),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_style(
        locations = cells_body(columns = "Sex"),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_options(table.border.top.style = "hidden") |>
    cols_label(
        Percentage = "Proportion (%)"    
    )
```

## Age and Sex

```{r tbl-sex-and-age}
#| include: true
#| tbl-cap: Sex of Australians as a proportion of each decade-wide age group.
#| label: tbl-sex-and-age
data |>
    mutate(age_group = case_when(
        age < 30                ~ "18-29",
        age >= 30 & age < 40    ~ "30-40",
        age >= 40 & age < 50    ~ "40-50",
        age >= 50 & age < 60    ~ "50-60",
        age >= 60 & age < 70    ~ "60-70",
        age >= 70               ~ "70+",
    )) |>
    select(-age, -total) |>
    group_by(age_group) |>
    summarise(across(male:female, sum)) |>
    pivot_longer(
        male:female,
        names_to = "sex",
        values_to = "total"
    ) |>
    group_by(age_group) |>
    mutate(percentage = specify_decimal(total / sum(total) * 100)) |>
    select(-total) |>
    pivot_wider(names_from = "sex", values_from = "percentage") |>
    relocate(female, .before = male) |>
    ungroup() |>
    gt() |>
    tab_spanner(
        label = "Proportion (%)",
        columns = c("female", "male"),
    ) |>
    tab_options(table.border.top.style = "hidden") |>
    cols_label(
        age_group = "Age",
        male = "Male",
        female = "Female"
    ) |>
    tab_style(
        locations = cells_body(columns = "age_group"),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_style(
        locations = cells_column_labels(),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_style(
        locations = cells_column_spanners(),
        style = list(
            cell_text(weight = "bold")
        )
    )
```


# Counts

```{r tbl-counts}
#| include: true
#| tbl-cap: Total number of Australians by age and sex.
#| label: tbl-counts

tbl_counts <- data |>
    relocate(female, .before = male) |>
    rename_with(str_to_sentence) |>
    gt() |>
    tab_style(
        locations = cells_column_labels(columns = everything()),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_style(
        locations = cells_body(columns = "Age"),
        style = list(
            cell_text(weight = "bold")
        )
    ) |>
    tab_options(table.border.top.style = "hidden") |>
    cols_label(Age = "Age (Years)")

tbl_counts
```